Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-25 Manuscript/manuscript.tex
Updates in methods section and comments added.

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig.~\ref{simVSexpNOions}) 
are proportional to the amount of bound charge
in a lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for $\alpha$ and $\beta$ segments in headgroup, as well as for
all C-H bonds in lipid molecules,
can be accurately measured by using H$^2$ NMR or $^{13}$C NMR techniques.
The experimental order parameters can be compared with the the ones
calculated from MD simulations by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

According to the electrometer concept,
the change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data, combined with the
information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}.

The experimental data used to quantify cation binding reports
order parameters as a function of equilibrium cation concentration
in the bulk solvent \cite{akutsu81,altenbach84}. Recent work
by NMRlipids project estimated this concentration by using
the relative molar concentration in water
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M,
which was accurate enough for the conclusions in that study \cite{catte16}.
Since we are now targeting a model with quantitatively accurate binding
affinity, we rather use a definition that corresponds more closely to the 
concentrations reported in experiments. 
By using larger simulation box dimensions that allow the
determination of bulk ion concentration, the equilibrium concentration
corresponding to the experimental data in the units of $\mathrm{mol/l}$
can be then determined from the simulations as
\begin{equation}
\label{eq:conc}
  C_{eq}=\frac{C_{bulk}}{0.602},
\end{equation}
where the bulk concentration, $C_{bulk}$, is the number density in the units of $\mathrm{nm}^{-3}$
in the bulk solvent far from the membrane interface. 
\todo{SAMULI: Once we have to final results, we can probably say that
  the repeat distance is not far from the experimentally measured distance \cite{pabst07,uhrikova08}
  }

The ion binding affinities between different models were compared by calculating
the relative surface excess of ions with respect to water, $\Gamma_i^w$, as \cite{chattorajBOOK}
\begin{eqnarray}
  \Gamma_i^w   &=& \Gamma_i-\Gamma_w \frac{C_i-C_i'}{C_w-C_w'} \\
  \Gamma_{a} &=& \frac{n_{a} - C_{a} V - C_{a}' V'}{A} \, \, ; \, a=i,w \, ,
\end{eqnarray}
where index $i$ and $w$ denote ions and water respectively,
prime accent denotes the phase inside bilayer, 
whereas non-prime symbols account for bulk water phase,
$n$ is the total amount of water or ions,
$C$ denotes bulk concentrations, 
$V$ is the volume of the respective phase separated by the Gibbs divinding plane, 
$A$ is the area of the interface, 
and $\Gamma_a$ is the surface excess of water or ions. 
As $\Gamma_i^w$ does not depend on the definition of the interfacial region
and the position of the Gibbs dividing plane~\cite{chattorajBOOK},
we consider the whole simulation box as an interface. 
Note that phospholipid membranes are bilayers, 
and hence the interface occurs in the simulation box twice
providing a surface area $2A$.
The concentrations of water and ions inside
the bilayer can be expected to vanish, thus $c_{i,w}'=0$.
This provides us a simplified relation for 
the relative surface excess of lipid bilayers in simulations
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where the bulk concentrations, $C_w$ and $C_i$, can be determined
from the simulation as $C_{bulk}$ in Equation~\ref{eq:conc}.

@@

Order parameters were calculated from simulations for all C-H bonds
by using Eq. \ref{OP} and form factors as
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

@@

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35 or 50 surfactants per leaflet.
These systems were first simulated for 200~ns using Lipid14 model.
First 20~ns was omitted from the analysis. 
Simulations with ECC-lipids used the same simulation setup.
We also applied electronic continuum correction to the cationic surfactant 
by scaling all its charges with the same factor as for ECC-lipids, $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.0cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
\end{figure}

\begin{table}
  \caption{Area per lipid (APL) from different models of POPC without ions\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids &        &  \\
    %Lipid14ecc0.80+sigma0.875 &        &  313    \\
    ($4.6\cdot 5.1 \, \mathrm{nm}^2$), 72 lipids patch, OPC3           & 63.2   &   313      \\
    \hline
    (6.4 nm)$^2$, 128 lipids patch, OPC3           & 64.2   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, SPC/E          & 65.1   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, OPC            & 64.4   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, TIP4p/2005     & 66.8   &  313       \\
    %($9.2\cdot 10.2 \, \mathrm{nm}^2$), 288 lipids patch           & 65.5   &  313       \\ %% not done for this model with f_sigma=0.89
    %oMM small patch           & 63.65  &         \\
    %oMM 4xbig patch           & 63.7   &         \\
    \hline
    %experiment   & 62.7  &  293    \\
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.} & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    %experiment  & 68.1  &  333    \\
    %experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of NaCl, CaCl$_2$ concentration
    from simulations with different force fields and experiments (DPPC \cite{akutsu81} and POPC \cite{altenbach84}).
    Simulations with Lipid14 parameters and \AA{}qvist ions are directly from Ref. \cite{catte16}, except that
    the concentration are determined with Eq. \ref{eq:conc}.
  }
  \todo{Simulation results with original Slipids and Dang ions to be added when we have the results.}
\end{figure*}

@@

Wide range of data presented in previous work \cite{catte16} suggested
that CaCl$_2$ binding behaviour cannot be corrected by only improving
ion models. This was not, however, tested specifically for Lipid14 model, so we
ran also Lipid14 simulations with ions by Dang~\cite{smith94,chang1999,dang2006} 
and ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014},
which have more realistic bulk behaviour than \AA{}qvist model~\cite{aqvist90}.
In line with previous work, the headgroup order parameters response
to CaCl$_2$ concentration was overestimated also with these ion models,
as seen in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively
\todo{Add OP-response of Lipid14+Dang in Fig. \ref{OrderParameterCHANGESnewMODELS}
  and OP-response of Lipid14+ECC-ions plot in SI}.

@@

>>>>

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and 
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of bound charge in a lipid bilayer \cite{seelig87}.
This is rationalized as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
More recent analysis in the NMRlipids project concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. The concept was also shown
to be suitable for the comparison of ion binding affinity in lipid 
bilayers between simulations and experiments~\cite{catte16}.

The order parameters for $\alpha$ and $\beta$ segments in headgroup, as well as for
all C-H bonds in lipid molecules,
can be accurately measured by using $^2$H NMR or $^{13}$C NMR techniques.
The order parameters can be calculated from MD simulations 
by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

According to the electrometer concept,
the change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
Before using electrometer concept to analyze cation binding, 
we calibrate the response of headgroup order parameters to the bound
charge in the simulation models. This is done by comparing 
the slope $m_i$ in Eq. \ref{OPchangeEQ} between simulations and
experiments of cationic surfactants in POPC bilayer \cite{scherer89}. 
Since essentially all surfactants locate in bilayer, the 
molar ratio directly gives the amount of bound charge per 
lipid $X^\pm$ is these systems.

\subsection{Ion concentrations and binding affinity}
The experimental headgroup order parameter data used
in this work is reported as a function of 
cation concentration remaining in the bulk 
solvent \cite{akutsu81,altenbach84}. 
To determine the same concentration from simulations, we
use large enough box to ensure that the ion concetration
reaches a clear plateau and the electroneutrality
is satisfied at the edge of simulation box. It should be noted that the recent work
by NMRlipids project compared simulations to the same 
experimental data by using the relative molar concentration in water
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M \cite{catte16}.
This is not exactly the same as the concentration reported in the used experimental
data, because the binding of ions in bilayer reduces the bulk concentration.
The difference slightly moves points towards lower concentrations but do not 
affect the conclusions in the previous study \cite{catte16}.

The ion binding affinities between different models were compared by calculating
the relative surface excess of ions with respect to water, $\Gamma_i^w$, as \cite{chattorajBOOK}
\begin{eqnarray}
  \Gamma_i^w   &=& \Gamma_i-\Gamma_w \frac{C_i-C_i'}{C_w-C_w'} \\
  \Gamma_{a} &=& \frac{n_{a} - C_{a} V - C_{a}' V'}{A} \, \, ; \, a=i,w \, ,
\end{eqnarray}
where index $i$ and $w$ denote ions and water respectively,
prime accent denotes the phase inside bilayer, 
whereas non-prime symbols account for bulk water phase,
$n$ is the total amount of water or ions,
$C$ denotes bulk concentrations, 
$V$ is the volume of the respective phase separated by the Gibbs divinding plane, 
$A$ is the area of the interface, 
and $\Gamma_a$ is the surface excess of water or ions. 
As $\Gamma_i^w$ does not depend on the definition of the interfacial region
and the position of the Gibbs dividing plane~\cite{chattorajBOOK},
we consider the whole simulation box as an interface. 
Note that phospholipid membranes are bilayers, 
and hence the interface occurs in the simulation box twice
providing a surface area $2A$.
The concentrations of water and ions inside
the bilayer can be expected to vanish, thus $c_{i,w}'=0$.
This provides us a simplified relation for 
the relative surface excess of lipid bilayers in simulations
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where the bulk concentrations, $C_w$ and $C_i$, can be determined
from the concentration plateau at the edge of the simulation box.

@@

The order parameters were calculated from simulations for all C-H bonds
in lipid molecule by using Eq. \ref{OP}. Form factors were calculated 
from equation
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

@@

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35 or 50 surfactants per leaflet.
The systems were simulated for 200~ns using Lipid14 model for POPC.
First 20~ns was omitted from the analysis.

The same systems were also simulated with ECC-lipid model for POPC \todo{How long?}.
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all \todo{Were all scaled? Are there charges is tails? Were those scaled as well in this case?} 
charges with the same factor as for ECC-lipids, $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.0cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{There seems to a typo in experimental value for other order parameter for C2 carbon in sn-2 chain; it should be negative.
  The scale of the y-axis should be then zoomed when this is fixed.} \\
  \todo{x-axis scale in form factor figure from 0 to 0.6, until where experimental data is available.}
\end{figure}

\begin{table}
  \caption{Area per lipid (APL) from different models of POPC without ions\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids &        &  \\
    %Lipid14ecc0.80+sigma0.875 &        &  313    \\
    ($4.6\cdot 5.1 \, \mathrm{nm}^2$), 72 lipids patch, OPC3           & 63.2   &   313      \\
    \hline
    (6.4 nm)$^2$, 128 lipids patch, OPC3           & 64.2   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, SPC/E          & 65.1   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, OPC            & 64.4   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, TIP4p/2005     & 66.8   &  313       \\
    %($9.2\cdot 10.2 \, \mathrm{nm}^2$), 288 lipids patch           & 65.5   &  313       \\ %% not done for this model with f_sigma=0.89
    %oMM small patch           & 63.65  &         \\
    %oMM 4xbig patch           & 63.7   &         \\
    \hline
    %experiment   & 62.7  &  293    \\
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.} & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    %experiment  & 68.1  &  333    \\
    %experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
  \todo{SAMULI: I would put here Lipid14 in 303K, ECC-lipid in 303K and experiment in 303K. Rest in SI.
  The best experimental value would be the one analyzed from the form factor shown in previous figure, 
  if available.}
\end{table}

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Changes of headgroup order parameters of POPC bilayer as a function of NaCl and CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the plateau of cation number density $C_{np}$
    at the box edge as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from Ref. \cite{catte16}.
  }
\end{figure*}

@@

Wide range of data presented in previous work \cite{catte16} suggested
that CaCl$_2$ binding behaviour cannot be corrected by only improving
ion models. This was not, however, tested specifically for Lipid14 model, so we
ran also Lipid14 simulations with ions by Dang~\cite{smith94,chang1999,dang2006} 
and ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014},
which have more realistic bulk behaviour than \AA{}qvist model~\cite{aqvist90}.
In line with previous work, the headgroup order parameters response
to CaCl$_2$ concentration was overestimated also with these ion models,
as seen in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.

@@

\todo{SAMULI: Maybe we should discuss the repeat distances and
  area per molecules measured at \cite{petrache06,pabst07,uhrikova08}
  }
